---
title: "Marianela Visualizations"
author: "Julia Martin"
date: "2026-06-10"
output:
    output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(tidyverse)
library(stringr)
library(paletteer)
library(plotly)
library(RColorBrewer)

```

## Altered vcd2df Function

```{r}
vcd2df_altered <- function(f_name, format) {
  # Basically, do some hacks to read line-by-line, and skip to the good part (vars)
  if (format != 'wide' && format != 'long') {
    stop("Please enter either 'long' or 'wide'")
  }
  
  fptr <- file(f_name, "r")
  line <- readLines(fptr, n = 1)
  
  # Read the variables and perform a few transposes
  # We need vars to rename the rows latter, but easier to use aliases atm.
  vars <- list()
  while (!grepl("\\$enddefinitions", line)) {
    if (!grepl("\\$scope", line)) {
      line <- readLines(fptr, n = 1)
    }
    if (grepl("var", line)) {
      # get the name and encoding and save it, if novel
      # we assume same named vars if different modules are redundant
      parts <- strsplit(line, " ")[[1]]
      if (!parts[4] %in% vars) {
        vars[[parts[5]]] <- parts[4]
      }
    }
    line <- readLines(fptr, n = 1)
  }
  # We make a df to store a time point
  t_n <- rep(-1, length(vars))
  names(t_n) <- vars
  df <- data.frame(t_n)
  
  repeat {
    line <- readLines(fptr, n = 1)
    if (length(line) == 0 || grepl("\\$dumpvars", line)) break
  }

  # Now we have the variables, we can read the dumpvars
  # We know there's a #0 somewhere, so...
  time <- '#0'
  while (TRUE) {
    while (length(line) > 0 && substring(line, 1, 1) != "#") {
      # lines either store a bit or a word, we parse differently
      if (grepl(" ", line)) { 
        # word
        parts <- strsplit(substring(line, 2), " ")[[1]]
        val <- parts[1]
        var <- trimws(parts[2])
      } else { 
        # bit
        val <- substring(line, 1, 1)
        var <- trimws(substring(line, 2))
      }
      # once parsed, verify its a unique register then store it
      if (var %in% names(t_n)) {
        # special case - just log non-numerics as -1
        # mostly happens during initialization
        if (grepl("^[0-9]+$", val)) {
          t_n[var] <- as.integer(strtoi(val, base = 2))
        } else {
          t_n[var] <- -1
        }
      }
      line <- readLines(fptr, n = 1)
    }
    # We have to update or single tick df and our result df, I guess
    df$vals <- t_n
    # We name the new column by the old time, then update the time
    names(df)[ncol(df)] <- time
    time <- line
    if (length(line) != 0) {
      # If we got another timestamp, we proceed.
      line <- readLines(fptr, n = 1)
    } else {
        close(fptr)
        rownames(df) <- names(vars)
        break
      }
  }
  
  if (format == 'wide'){
    return(df[, -1])
    }
  
  else if (format == 'long') {
  long_df = df[, -1] %>%
    rownames_to_column() %>%
    gather(colname, value, -rowname)
  names(long_df) <- c("register", "timestamp", "value")
  return(long_df)
  }
}
```



# Figure 1

## read in data

```{r}
test_1 = vcd2df_altered("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds/alu_add_sub.vcd", "long")
```

## wrangle data

```{r}
test_1_viz = test_1 %>%
  filter(!grepl("^shadow_", register)) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/n()) %>%
  distinct(prefix, percent) %>%
  mutate(prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))
```

## visualize data

```{r}
test_1_viz %>% 
  ggplot(aes(x = prefix, y = percent, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Figure 1: Usage of Register Pathways for All Timestamps",
    subtitle = "Data from .vcd file 'alu_add_sub'",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```

# Figure 2

## read in data

```{r}
files_four <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds",
  full.names = TRUE
)

vcd_folder <- lapply(files_four, vcd2df_altered, "long")
names(vcd_folder) <- tools::file_path_sans_ext(basename(files_four))
list2env(vcd_folder, envir = .GlobalEnv)
```

## wrangle data

```{r}
four_vcd_df <- bind_rows(alu_add_sub, alu_eq, alu_lts, alu_ltu)

four_vcd_df_viz = four_vcd_df %>%
  filter(!grepl("^shadow_", register)) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/n()) %>%
  distinct(prefix, percent) %>%
  mutate(prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))

four_vcd_df_viz
```

## visualize data

```{r}
four_vcd_df_viz %>% 
  ggplot(aes(x = prefix, y = percent, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Figure 2: Usage of Register Pathways for All Timestamps",
    subtitle = "Data from 4 .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```

# Figure 1.1

## read in data

```{r}
test_1 = vcd2df_altered("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds/alu_add_sub.vcd", "long")
```

## wrangle data

```{r}
test_1_viz = test_1 %>%
  filter(!grepl("^shadow_", register)) %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity)) %>%
  distinct(prefix, mean)

test_1_viz %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Figure 1.1: Usage of Active Register Pathways for All Timestamps",
    subtitle = "Data from .vcd file 'alu_add_sub'",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```

# Figure 2.1

## read in data

```{r}
files_four <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds",
  full.names = TRUE
)

vcd_folder <- lapply(files_four, vcd2df_altered, "long")
names(vcd_folder) <- tools::file_path_sans_ext(basename(files_four))
list2env(vcd_folder, envir = .GlobalEnv)
```

## wrangle data

```{r}
four_vcd_df <- bind_rows(alu_add_sub, alu_eq, alu_lts, alu_ltu)

four_vcd_df_viz = four_vcd_df %>%
  filter(!grepl("^shadow_", register)) %>%
  mutate(n_timestamp = n_distinct(timestamp) * 4) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity)) %>%
  distinct(prefix, mean)
```

## visualize data

```{r}
four_vcd_df_viz %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Figure 2.1: Usage of Register Pathways for All Timestamps (4 .vcd files)",
    subtitle = "Data from 4 .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```

# Figure 3

## read in data

```{r}
files_four <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds",
  full.names = TRUE
)

vcd_folder <- lapply(files_four, vcd2df_altered, "long")
names(vcd_folder) <- tools::file_path_sans_ext(basename(files_four))
list2env(vcd_folder, envir = .GlobalEnv)
```


## wrangle data

```{r}
four_vcd_df <- bind_rows(alu_add_sub, alu_eq, alu_lts, alu_ltu)

df_viz <- four_vcd_df %>%
  filter(!grepl("^shadow_", register)) %>%
  mutate(n_timestamp = n_distinct(timestamp) * 4) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  summarise(
    mean = mean(activity),
    n_registers = n_distinct(register),
    min_activity = min(activity),
    max_activity = max(activity),
    median_activity = median(activity),
    .groups = "drop"
  )
```


## visualize data

```{r}
p <- ggplot(df_viz, aes(x = prefix, y = mean, fill = prefix,
           text = paste0(
             "Prefix: ", prefix, "\n",
             "# Registers: ", n_registers, "\n",
             "Min Activity: ", round(min_activity, 4), "\n",
             "Max Activity: ", round(max_activity, 4), "\n",
             "Median Activity: ", round(median_activity, 4)))) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(title = "Figure 3: Usage of Register Pathways for All Timestamps",
         subtitle = "Data from .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
         x = "Register Prefix",
         y = "Percent of Total Usage")

ggplotly(p, tooltip = "text") %>%
  layout(
    annotations = list(
      list(
        text = "Data from .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
        x = 0,
        y = 1.035,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left"
      )
    )
  )
```

# Figure 4

## read in data

```{r}
test_1 = vcd2df_altered("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds/alu_add_sub.vcd", "long")
```

## wrangle data

```{r}
test_1_viz <- test_1 %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  filter(!grepl("^shadow_", register)) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp,
         prefix = sub("_.*", "", register)) %>%
  ungroup() %>%
  group_by(prefix) %>%
  summarise(
    mean = mean(activity),
    n_registers = n_distinct(register),
    min_activity = min(activity),
    max_activity = max(activity),
    median_activity = median(activity),
    .groups = "drop")
```

## visualize data

```{r}
p <- ggplot(test_1_viz, aes(x = prefix, y = mean, fill = n_registers,
           text = paste0(
             "Prefix: ", prefix, "\n",
             "# sub-Registers: ", n_registers, "\n",
             "Min Activity: ", round(min_activity, 4), "\n",
             "Max Activity: ", round(max_activity, 4), "\n",
             "Median Activity: ", round(median_activity, 4)))) +
    geom_col() +
    scale_fill_gradient(low = "darkgrey", high = "darkred") +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "right") +
    labs(title = "Figure 4: Usage of Register Pathways for All Timestamps",
         subtitle = "Data from .vcd files 'alu_add_sub'",
         x = "Register Prefix",
         y = "Percent of Total Usage")

ggplotly(p, tooltip = "text") %>%
  layout(
    annotations = list(
      list(
        text = "Data from .vcd file 'alu_add_sub'",
        x = 0,
        y = 1.035,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left"
      )
    )
  )
```

# Figure 5

## create graphing function

```{r}
prefix_3d_viz_2vcd <- function(file_location) {
  
  # make sure the vcd2df file is loaded
  if (!exists("vcd2df_altered", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_altered function is loaded into your environment.")
  }
  
  # turn the folder of .vcd files into .df 
  files <- list.files(file_location, full.names = TRUE)
  
  # check that the three required columns are present for all files
  required_cols <- c("register", "timestamp", "value")
  
  # create a function that processes each individual file
  process_file <- function(file) {

    df <- vcd2df_altered(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    
    ## within this fn create the dfs (one for % usage of all registers which is used to find mean usage for prefix registers)
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      summarise(mean = mean(activity),
                std = ifelse(n() > 1, sd(activity), 0),
                n = n(),
                .groups = "drop") %>%
      filter(std > 0.1)

    return(df_viz)
  }
  
  results <- lapply(files, process_file)
  
  names(results) <- tools::file_path_sans_ext(basename(files))
  
  combined <- bind_rows(results, .id = "source")

  ### now graph the data
  
  plot_ly(
    combined,
    x = ~seq_along(prefix),
    y = ~mean,
    z = ~std,
    type = "scatter3d",
    mode = "markers+text",
    text = ~prefix,
    color = ~source
) %>%
  layout(
    title = list(
      text = "Figure 5: 3-axis Scatterplot",
      y = 0.95,
      x = 0.5,
      xanchor = "center",
      yanchor = "top"
    ),
  scene = list(
    xaxis = list(title = "Prefix"),
    yaxis = list(title = "Std. Deviation"),
    zaxis = list(title = "Mean"),
    bgcolor = "lightgrey"
    )
  )
  
  }
```


## visualize data

```{r}
prefix_3d_viz_2vcd("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/2_vcds")
```



# Figure 6

## create graphing function

```{r}
subgroup_3d_viz_all_vcd <- function(file_location) {
  
  # make sure the vcd2df file is loaded
  if (!exists("vcd2df_altered", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_altered function is loaded into your environment.")
  }
  
  # turn the folder of .vcd files into .df 
  files <- list.files(file_location, full.names = TRUE)
  
  # check that the three required columns are present for all files
  required_cols <- c("register", "timestamp", "value")
  
  # create a function that processes each individual file
  process_file <- function(file) {

    df <- vcd2df_altered(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    
    ## within this fn create the dfs (one for % usage of all registers which is used to find mean usage for prefix registers)
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      mutate(mean = mean(activity),
             std = ifelse(n() > 1, round(sd(activity), 4), 0),
             n = n(),
             subgroup = sub(".*_([^_]+).*", "\\1", register),
             activity = round(activity, 4)) %>%
      distinct(prefix, activity, subgroup, std) %>%
      filter(std > 0.1)

    return(df_viz)
  }
  
  results <- lapply(files, process_file)
  
  names(results) <- tools::file_path_sans_ext(basename(files))
  
  combined <- bind_rows(results, .id = "source")

  ### now graph the data
  
  # identify structure
  prefix_levels <- sort(unique(combined$prefix))
  source_levels <- unique(combined$source)
  
  n_prefix <- length(prefix_levels)
  n_source <- length(source_levels)
  
  # generate palettes for each source automatically
  palette_list <- brewer.pal(min(max(n_source, 3), 8), "Set1")
  if (n_source > 8) {
    palette_list <- colorRampPalette(palette_list)(n_source)
  }
  
  # map source -> base palette
  source_colors <- setNames(palette_list, source_levels)
  
  combined <- combined %>%
    mutate(
      prefix_id = match(prefix, prefix_levels)
    )
  
  combined$color <- mapply(
    function(src, pid) {
  
      base_col <- source_colors[[src]]
  
      # create light-to-dark variation per prefix
      ramp <- colorRampPalette(c("white", base_col, "black"))
  
      ramp(n_prefix)[pid]
    },
    combined$source,
    combined$prefix_id
  )

  
  plot_ly(
    combined,
    x = ~seq_along(prefix),
    y = ~activity,
    z = ~std,
    type = "scatter3d",
    mode = "markers",
    color = ~I(color),
    symbol = ~source,
    text = ~paste(
      "Dataset:", source,
      "<br>Prefix:", prefix,
      "<br>Activity:", activity,
      "<br>Std:", std),
    hoverinfo = "text",
    marker = list(opacity = 0.8)) %>%
    layout(
    title = list(
      text = "Figure 6: 3-axis Scatterplot Subgroups",
      y = 0.95,
      x = 0.5,
      xanchor = "center",
      yanchor = "top"
    ),
  scene = list(
    xaxis = list(title = "Prefix"),
    yaxis = list(title = "Std. Deviation"),
    zaxis = list(title = "Mean"),
    bgcolor = "lightgrey"
      )
    )
  }
```


## visualize data

```{r}
subgroup_3d_viz_all_vcd("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds")
```

