---
title: "Actual RMD File"
author: "Julia Martin"
date: "2026-06-10"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(reticulate)
use_python("C:/Users/marin/AppData/Local/Programs/Python/Python312/python.exe", required = TRUE)

library(tidyverse)
library(stringr)
library(paletteer)
library(plotly)
library(RColorBrewer)

#library(vcd2df)
```

# Functions

## original function

```{r}
vcd2df <- function(f_name) {
  # Basically, do some hacks to read line-by-line, and skip to the good part (vars)
  fptr <- file(f_name, "r")
  line <- readLines(fptr, n = 1)
  while (!grepl("\\$scope", line)) {
      line <- readLines(fptr, n = 1)
  }

  # Read the variables and perform a few transposes
  # We need vars to rename the rows latter, but easier to use aliases atm.
  vars <- list()
  while (!grepl("\\$enddefinitions", line)) {
    if (grepl("var", line)) {
      # get the name and encoding and save it, if novel
      # we assume same named vars if different modules are redundant
      parts <- strsplit(line, " ")[[1]]
      if (!parts[4] %in% vars) {
        vars[[parts[5]]] <- parts[4]
      }
    }
    line <- readLines(fptr, n = 1)
  }
  # We make a df to store a time point
  t_n <- rep(-1, length(vars))
  names(t_n) <- vars
  df <- data.frame(t_n)
  
  while (!grepl("\\$dumpvars", readLines(fptr, n = 1))) {
    # Keep reading until $dumpvars is found
  }

  # Now we have the variables, we can read the dumpvars
  # We know there's a #0 somewhere, so...
  time <- '#0'
  while (TRUE) {
    while (length(line) > 0 && substring(line, 1, 1) != "#") {
      # lines either store a bit or a word, we parse differently
      if (grepl(" ", line)) { 
        # word
        parts <- strsplit(substring(line, 2), " ")[[1]]
        val <- parts[1]
        var <- trimws(parts[2])
      } else { 
        # bit
        val <- substring(line, 1, 1)
        var <- trimws(substring(line, 2))
      }
      # once parsed, verify its a unique register then store it
      if (var %in% names(t_n)) {
        # special case - just log non-numerics as -1
        # mostly happens during initialization
        if (grepl("^[0-9]+$", val)) {
          t_n[var] <- as.integer(strtoi(val, base = 2))
        } else {
          t_n[var] <- -1
        }
      }
      line <- readLines(fptr, n = 1)
    }
    # We have to update or single tick df and our result df, I guess
    df$vals <- t_n
    # We name the new column by the old time, then update the time
    names(df)[ncol(df)] <- time
    time <- line
    if (length(line) != 0) {
      # If we got another timestamp, we proceed.
      line <- readLines(fptr, n = 1)
    } else {
      # Or we change from encodings back to reg names, then return
      close(fptr)
      rownames(df) <- names(vars)
      return(df[, -1])
    }
  }
}
```

## original function altered by me

```{r}
vcd2df_test <- function(f_name, format) {
  # Basically, do some hacks to read line-by-line, and skip to the good part (vars)
  if (format != 'wide' && format != 'long') {
    stop("Please enter either 'long' or 'wide'")
  }
  
  fptr <- file(f_name, "r")
  line <- readLines(fptr, n = 1)
  
  # Read the variables and perform a few transposes
  # We need vars to rename the rows latter, but easier to use aliases atm.
  vars <- list()
  while (!grepl("\\$enddefinitions", line)) {
    if (!grepl("\\$scope", line)) {
      line <- readLines(fptr, n = 1)
    }
    if (grepl("var", line)) {
      # get the name and encoding and save it, if novel
      # we assume same named vars if different modules are redundant
      parts <- strsplit(line, " ")[[1]]
      if (!parts[4] %in% vars) {
        vars[[parts[5]]] <- parts[4]
      }
    }
    line <- readLines(fptr, n = 1)
  }
  # We make a df to store a time point
  t_n <- rep(-1, length(vars))
  names(t_n) <- vars
  df <- data.frame(t_n)
  
  repeat {
    line <- readLines(fptr, n = 1)
    if (length(line) == 0 || grepl("\\$dumpvars", line)) break
  }

  # Now we have the variables, we can read the dumpvars
  # We know there's a #0 somewhere, so...
  time <- '#0'
  while (TRUE) {
    while (length(line) > 0 && substring(line, 1, 1) != "#") {
      # lines either store a bit or a word, we parse differently
      if (grepl(" ", line)) { 
        # word
        parts <- strsplit(substring(line, 2), " ")[[1]]
        val <- parts[1]
        var <- trimws(parts[2])
      } else { 
        # bit
        val <- substring(line, 1, 1)
        var <- trimws(substring(line, 2))
      }
      # once parsed, verify its a unique register then store it
      if (var %in% names(t_n)) {
        # special case - just log non-numerics as -1
        # mostly happens during initialization
        if (grepl("^[0-9]+$", val)) {
          t_n[var] <- as.integer(strtoi(val, base = 2))
        } else {
          t_n[var] <- -1
        }
      }
      line <- readLines(fptr, n = 1)
    }
    # We have to update or single tick df and our result df, I guess
    df$vals <- t_n
    # We name the new column by the old time, then update the time
    names(df)[ncol(df)] <- time
    time <- line
    if (length(line) != 0) {
      # If we got another timestamp, we proceed.
      line <- readLines(fptr, n = 1)
    } else {
        close(fptr)
        rownames(df) <- names(vars)
        break
      }
  }
  
  if (format == 'wide'){
    return(df[, -1])
    }
  
  else if (format == 'long') {
  long_df = df[, -1] %>%
    rownames_to_column() %>%
    gather(colname, value, -rowname)
  names(long_df) <- c("register", "timestamp", "value")
  return(long_df)
  }
}
```

## % of signal transactions

```{r}
prefix_change_summary <- function(df) {

  df %>%
    mutate(prefix = sub("_.*", "", register)) %>%
    group_by(register) %>%
    summarise(changes = sum(value != lag(value), na.rm = TRUE),
              prefix = first(prefix), 
              .groups = "drop") %>%
    group_by(prefix) %>%
    summarise(changes = sum(changes), 
              .groups = "drop") %>%
    mutate(percent = changes / sum(changes)) %>%
    arrange(desc(percent))
  
  }
```

## vcd_viz_file (1 file)

```{r}
vcd_viz_file <- function(df, file_location) {
  
  # check whether the df is in long format and whether the columns are named correctly
  required_cols <- c("register", "timestamp", "value")

  if (!all(required_cols %in% names(df))) {
    stop("Please make sure columns are titled 'register', 'timestamp', and 'value' accordingly.")
  }

  # check whether ggplot is loaded
  
  if (!requireNamespace("tidyverse", quietly = TRUE)) {
  stop("Please make sure ggplot2 is installed and available.")
}
  
  # get the # of total files so we can multiply timestamp by the correct amount (necessary to have the percentage be calculated correctly)
  num_files = length(list.files(file_location))
  
  df_viz = df %>%
    mutate(n_timestamp = n_distinct(timestamp) * num_files) %>%
    filter(value != 0) %>%
    group_by(register) %>%
    mutate(n_value_change = n(), 
           activity = n_value_change / n_timestamp) %>%
    distinct(register, activity) %>%
    arrange(register)

  df_mean = df_viz %>%
    mutate(prefix = sub("_.*", "", register)) %>%
    group_by(prefix) %>%
    mutate(mean = mean(activity)) %>%
    distinct(prefix, mean)
  
  df_mean2 <- df_mean %>%
    mutate(plot_type = "Prefix Registers")

  df_viz2 <- df_viz %>%
    mutate(plot_type = "All Registers")
  
  df_combined <- bind_rows(
    df_mean2 %>% select(x = prefix, y = mean, plot_type),
    df_viz2 %>% select(x = register, y = activity, plot_type)
  )
  
  df_combined_viz = df_combined %>%
    ggplot(aes(x = x, y = y, fill = x)) +
    geom_col() +
    facet_grid(. ~ plot_type, scales = "free_x") +
    scale_y_continuous(labels = scales::percent) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
      legend.position = "none"
    )
  
  return(df_combined_viz)
  
  }
```

## vcd_viz_folder (folder)

```{r}
vcd_viz_folder <- function(file_location) {
  
  # make sure the vcd2df file is loaded
  if (!exists("vcd2df_test", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_test function is loaded into your environment.")
  }
  
  # turn the folder of .vcd files into .df 
  files <- list.files(file_location, full.names = TRUE)
  
  # check that the three required columns are present for all files
  required_cols <- c("register", "timestamp", "value")
  
  # create a function that processes each individual file
  process_file <- function(file) {

    df <- vcd2df_test(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    
    ## within this fn create the dfs (one for % usage of all registers which is used to find mean usage for prefix registers)
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      arrange(register)
      # distinct(register, activity)
    
    df_mean <- df_viz %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      mutate(mean = mean(activity)) %>%
      distinct(prefix, mean)


    return(df_mean)
  }
  
  results <- lapply(files, process_file)
  
  names(results) <- tools::file_path_sans_ext(basename(files))
  
  df_mean_all <- bind_rows(results, .id = "plot_type")

  ggplot(df_mean_all, aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    facet_wrap(~ plot_type, scales = "free_x") +
    scale_y_continuous(labels = scales::percent) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
      legend.position = "none"
    )
  
}
```


## vcd_viz_folder_interactive

```{r}
vcd_viz_folder_int <- function(file_location) {
  
  if (!exists("vcd2df_test", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_test function is loaded into your environment.")
  }
  
  files <- list.files(file_location, full.names = TRUE)
  required_cols <- c("register", "timestamp", "value")
  
  process_file <- function(file, file_name) {
    df <- vcd2df_test(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      arrange(register)
    
    df_mean <- df_viz %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      summarise(
        mean = mean(activity),
        n_registers = n_distinct(register),
        min_activity = min(activity),
        max_activity = max(activity),
        median_activity = median(activity),
        .groups = "drop"
      ) %>%
      mutate(plot_type = file_name)
    
    return(df_mean)
  }
  
  # Get clean file names
  file_names <- tools::file_path_sans_ext(basename(files))
  
  # Use Map to pass both file path and file name
  results <- Map(process_file, files, file_names)
  
  df_mean_all <- bind_rows(results)
  
  p <- ggplot(df_mean_all, aes(x = prefix, y = mean, fill = prefix,
                               text = paste0(
                                 "Prefix: ", prefix, "\n",
                                 "# Registers: ", n_registers, "\n",
                                 "Min Activity: ", round(min_activity, 4), "\n",
                                 "Max Activity: ", round(max_activity, 4), "\n",
                                 "Median Activity: ", round(median_activity, 4)
                               ))) +
    geom_col() +
    facet_wrap(~ plot_type, scales = "free_x") +
    scale_y_continuous(labels = scales::percent) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
      legend.position = "none"
    )
  
  plotly::ggplotly(p, tooltip = "text")
}
```


## vcd_viz_folder_interactive_shadow

```{r}
# oh god I have to comment throughout this
# i forgot to when writing it
# kill me

vcd_viz_folder_int_shadow <- function(file_location, type) {
  
  #check that type is either prefix or all, and return an error if not (necessary since my code relies on using those specific key words)
  if (type != 'all' && type != 'prefix') {
    stop("Please enter either 'all' (all registers) or 'prefix' (mean for each prefix)")
  }
  
  #make sure the vcd2df fn is loaded into the environment, as the vcd files can't be converted without it
  if (!exists("vcd2df_test", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_test function is loaded into your environment.")
  }
  
  # read files into memory(?) fn environment(?)
  files <- list.files(file_location, full.names = TRUE)
  
  # necessary for later, since we need to make sure the three columns are present for my code to work
  required_cols <- c("register", "timestamp", "value")
  
  ### define the fn take takes in a single file and does the necessary changes to make it graphable (more info in comments in fn)
  process_file <- function(file, file_name) {
    
    # run raw vcd file through the vcd2df function (using 'long' format is super important!!! don't change you'll have to completely rewrite fn!!!)
    df <- vcd2df_test(file, "long")
  
    # check that the file contains all three 'required_cols' defined (8 lines?) before
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    # creates % activity across all timestamps for each *individual* register 
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      filter(grepl("^shadow_", register)) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      arrange(register) %>%
      mutate(plot_type = file_name) # important to be graphable for some reason!! don't change!! took me way too long to figure this out!!
    
    # uses above df to group register subgroups '..._subgroup_...' and then calculate the mean, min, max, and median for the % activity 
    df_prefix <- df_viz %>%
      mutate(prefix = sub(".*_([^_]+)_.*", "\\1", register)) %>%
      group_by(prefix) %>%
      summarise(
        mean = mean(activity),
        n_registers = n_distinct(register),
        min_activity = min(activity),
        max_activity = max(activity),
        median_activity = median(activity),
        .groups = "drop"
      ) %>%
      mutate(plot_type = file_name) # again don't change or remove!!
    
    # checks whether user wanted prefix or all and returns that one
    if (type == 'prefix') {return(df_prefix)}
    else if (type == 'all') {return(df_viz)}
    
  }
  
  # Get clean file names (so the file name displays on graph)
  file_names <- tools::file_path_sans_ext(basename(files))
  
  # Use Map to pass both file path and file name
  results <- Map(process_file, files, file_names)
  
  # turn the returned dfs into one file with a column specifying which file it's for
  df <- bind_rows(results)
  
  # check if "all" was chosen since 'register' would be the colname and then graphs
  if ("register" %in% names(df)) {

    ggplot(df, aes(x = register, y = activity, fill = register)) +
    geom_col() +
    facet_wrap(~ plot_type, scales = "free_x") +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
          legend.position = "none")

  }

  # checks if "prefix" was chosen since "prefix" would be the colname and then does lots of fancy graphing (lwk not sure how this works if im being real)
  else if ("prefix" %in% names(df)) {

    # plotly works by taking in two things, a visualization and 'text' that's defined within the visualization and is generated by things in the graph
    # so 'p' is the graph, and 'text' is what will be displayed when hovering over the graph
    # if you want to alter what is shown when hovering over the graph, you need to change what is listed in the 'text = paste0(...)
    p <- ggplot(df, aes(x = prefix, y = mean, fill = n_registers,
                         text = paste0(
                           "Prefix: ", prefix, "\n",
                           "# Registers: ", n_registers, "\n",
                           "Min Activity: ", round(min_activity, 4), "\n",
                           "Max Activity: ", round(max_activity, 4), "\n",
                           "Median Activity: ", round(median_activity, 4)))) +
    geom_col() +
    scale_fill_gradient(low = "darkgrey", high = "darkred") +
    facet_wrap(~ plot_type, scales = "free_x") +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
          legend.position = "right")

  # this is just putting the graph 'p' and the tooltip "text" together
  plotly::ggplotly(p, tooltip = "text")

  }
}
```

## prefix_3d_viz_2vcd

```{r}
prefix_3d_viz_2vcd <- function(file_location) {
  
  # make sure the vcd2df file is loaded
  if (!exists("vcd2df_test", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_test function is loaded into your environment.")
  }
  
  # turn the folder of .vcd files into .df 
  files <- list.files(file_location, full.names = TRUE)
  
  # check that the three required columns are present for all files
  required_cols <- c("register", "timestamp", "value")
  
  # create a function that processes each individual file
  process_file <- function(file) {

    df <- vcd2df_test(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    
    ## within this fn create the dfs (one for % usage of all registers which is used to find mean usage for prefix registers)
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      summarise(mean = mean(activity),
                std = ifelse(n() > 1, sd(activity), 0),
                n = n(),
                .groups = "drop") %>%
      filter(std > 0.1)

    return(df_viz)
  }
  
  results <- lapply(files, process_file)
  
  names(results) <- tools::file_path_sans_ext(basename(files))
  
  combined <- bind_rows(results, .id = "source")

  ### now graph the data
  
  plot_ly(
    combined,
    x = ~seq_along(prefix),
    y = ~mean,
    z = ~std,
    type = "scatter3d",
    mode = "markers+text",
    text = ~prefix,
    color = ~source
) %>%
  layout(title = "vcd_1 vs vcd_2")
  
  }
```

## subgroup_3d_viz_all_vcd

```{r}
subgroup_3d_viz_all_vcd <- function(file_location) {
  
  # make sure the vcd2df file is loaded
  if (!exists("vcd2df_test", envir = .GlobalEnv)) {
    stop("Please make sure the vcd2df_test function is loaded into your environment.")
  }
  
  # turn the folder of .vcd files into .df 
  files <- list.files(file_location, full.names = TRUE)
  
  # check that the three required columns are present for all files
  required_cols <- c("register", "timestamp", "value")
  
  # create a function that processes each individual file
  process_file <- function(file) {

    df <- vcd2df_test(file, "long")
  
    if (!all(required_cols %in% names(df))) {
        stop("One file is missing required columns: register, timestamp, value")
    }
    
    
    ## within this fn create the dfs (one for % usage of all registers which is used to find mean usage for prefix registers)
    df_viz <- df %>%
      mutate(n_timestamp = n_distinct(timestamp)) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      mutate(mean = mean(activity),
             std = ifelse(n() > 1, round(sd(activity), 4), 0),
             n = n(),
             subgroup = sub(".*_([^_]+).*", "\\1", register),
             activity = round(activity, 4)) %>%
      distinct(prefix, activity, subgroup, std) %>%
      filter(std > 0.1)

    return(df_viz)
  }
  
  results <- lapply(files, process_file)
  
  names(results) <- tools::file_path_sans_ext(basename(files))
  
  combined <- bind_rows(results, .id = "source")

  ### now graph the data
  
  # identify structure
  prefix_levels <- sort(unique(combined$prefix))
  source_levels <- unique(combined$source)
  
  n_prefix <- length(prefix_levels)
  n_source <- length(source_levels)
  
  # generate palettes for each source automatically
  palette_list <- brewer.pal(min(max(n_source, 3), 8), "Set1")
  if (n_source > 8) {
    palette_list <- colorRampPalette(palette_list)(n_source)
  }
  
  # map source -> base palette
  source_colors <- setNames(palette_list, source_levels)
  
  combined <- combined %>%
    mutate(
      prefix_id = match(prefix, prefix_levels)
    )
  
  combined$color <- mapply(
    function(src, pid) {
  
      base_col <- source_colors[[src]]
  
      # create light-to-dark variation per prefix
      ramp <- colorRampPalette(c("white", base_col, "black"))
  
      ramp(n_prefix)[pid]
    },
    combined$source,
    combined$prefix_id
  )

  
  plot_ly(
    combined,
    x = ~seq_along(prefix),
    y = ~activity,
    z = ~std,
    type = "scatter3d",
    mode = "markers",
    color = ~I(color),
    symbol = ~source,
    text = ~paste(
      "Dataset:", source,
      "<br>Prefix:", prefix,
      "<br>Activity:", activity,
      "<br>Std:", std),
    hoverinfo = "text",
    marker = list(opacity = 0.8)) %>%
    layout(title = "subgroup all vcds",
           scene = list(
              bgcolor = "lightgrey"
    ))
  
  }
```

# USING ORIGINAL FN ALTERED BY ME

## Single Visualization

### read in data

```{r}
test_1 = vcd2df_test("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds/alu_add_sub.vcd", "long")
```

### prep data

```{r}
test_1_viz = test_1 %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/n()) %>%
  distinct(prefix, percent) %>%
  mutate(prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))
```

### visualize data

```{r}
test_1_viz %>% 
  ggplot(aes(x = prefix, y = percent, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Figure 1: Usage of Register Pathways for All Timestamps",
    subtitle = "Data from .vcd file 'alu_add_sub'",
    x = "Register Prefix",
    y = "Percent of Total Appearance"
  )
```


## 4 .vcd Files Visualization

### read in data

```{r}
files_four <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds",
  full.names = TRUE
)

vcd_folder <- lapply(files_four, vcd2df_test, "long")
names(vcd_folder) <- tools::file_path_sans_ext(basename(files_four))
list2env(vcd_folder, envir = .GlobalEnv)
```

### prep data for vizualization

```{r}
four_vcd_df <- bind_rows(alu_add_sub, alu_eq, alu_lts, alu_ltu)

four_vcd_df_viz = four_vcd_df %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/n()) %>%
  distinct(prefix, percent) %>%
  mutate(prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))
```

### visualize data

```{r}
four_vcd_df_viz %>% 
  ggplot(aes(x = prefix, y = percent, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Usage of Register Pathways for All Timestamps (4 .vcd files)",
    subtitle = "Data from 4 .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```


## whole .vcd folder Visualization

### read in data

```{r}
files_full <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds",
  full.names = TRUE
)

vcd_folder <- lapply(files_full, vcd2df_test, "long")
names(vcd_folder) <- tools::file_path_sans_ext(basename(files_full))
folder_vcd_df <- bind_rows(vcd_folder)
```

### prep data for vizualization

```{r}
folder_vcd_df_viz = folder_vcd_df %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  mutate(percent = n/n()) %>%
  distinct(prefix, percent) %>%
  mutate(prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))
```

### visualize data

```{r}
folder_vcd_df_viz %>% 
  ggplot(aes(x = prefix, y = percent, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
  labs(
    title = "Usage of Register Pathways for All Timestamps",
    subtitle = "Data from 4 .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```



# bleh

NOTES:
 * -1 might be something I filter out???
 -- keep it for now, but I'll ask Calvin

## testing

```{r, echo = FALSE}
sapply(vcd_folder, nrow)
sapply(vcd_folder, function(x) digest::digest(x))


four_vcd_df_viz = four_vcd_df %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(timestamp) %>%
  mutate(changes = sum(value != lag(value), na.rm = TRUE)) %>%
  group_by(prefix) %>%
  summarise(changes = sum(changes)) %>%
  mutate(percent = changes / sum(changes),
         prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))


test_1_viz = test_1 %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(register) %>%
  summarise(changes = sum(value != lag(value), na.rm = TRUE),
            prefix = first(prefix)) %>%
  group_by(prefix) %>%
  summarise(changes = sum(changes)) %>%
  mutate(percent = changes / sum(changes),
         prefix = forcats::fct_reorder(prefix, percent, .desc = TRUE))


n_distinct(four_vcd_df$timestamp)
n_distinct(four_vcd_df$register)
nrow(four_vcd_df)

```

## IDEA

 * Lucy graph ideas:
  - be able to hover over bars and show exact percentages


### one vcd

```{r}
test_1_viz = test_1 %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  arrange(register)

test_1_viz %>% 
  ggplot(aes(x = register, y = activity, fill = register)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

test_1_viz_mean = test_1_viz %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity)) %>%
  distinct(prefix, mean)

test_1_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none") +
  labs(
    title = "Figure 1.1: Usage of Active Register Pathways for All Timestamps",
    subtitle = "Data from .vcd file 'alu_add_sub'",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )
```


### four vcd

```{r}
#activity = (# times value changed) / (# timestamps)

four_vcd_df_viz = four_vcd_df %>%
  mutate(n_timestamp = n_distinct(timestamp) * 4) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  arrange(register)

four_vcd_df_viz %>% 
  ggplot(aes(x = register, y = activity, fill = register)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

four_vcd_df_viz_mean = four_vcd_df_viz %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity)) %>%
  distinct(prefix, mean)

four_vcd_df_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none") +
  labs(
    title = "Figure 2.1: Usage of Register Pathways for All Timestamps (4 .vcd files)",
    subtitle = "Data from 4 .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage"
  )

# okay so there are technically 4 repeated registers per timestamp due to combining the data 
# I guess I could just multiply the timestamp by 4?
# is that totally cheating????
```

### whole .vcd folder

```{r}
num_files = length(list.files("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds"))

folder_vcd_df_viz = folder_vcd_df %>%
  mutate(n_timestamp = n_distinct(timestamp) * num_files) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  arrange(register)

folder_vcd_df_viz %>% 
  ggplot(aes(x = register, y = activity, fill = register)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

folder_vcd_df_viz_mean = folder_vcd_df_viz %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity)) %>%
  distinct(prefix, mean)

folder_vcd_df_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")
```

### AHHHHSDFSDKFHSLDKFJ

```{r}
# OH MY GOD THEY'RE DIFFERENT
# BUT LIKE NOT THAT DIFFERENT?????
# BUT STILL DIFFERENT IN VERY VERY SMALL AMOUNTS OF YOU COMPARE THE MEANS

## separate graphs

test_1_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

four_vcd_df_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

folder_vcd_df_viz_mean %>% 
  ggplot(aes(x = prefix, y = mean, fill = prefix)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=5),
          legend.position = "none")

## stacked graph

combined_viz <- bind_rows(
  test_1_viz_mean %>% mutate(source = "test_1"),
  four_vcd_df_viz_mean %>% mutate(source = "four_vcd"),
  folder_vcd_df_viz_mean %>% mutate(source = "folder_vcd"))

combined_viz %>% 
  ggplot(aes(x = prefix, y = mean, fill = source)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
        legend.position = "right") +
  scale_fill_paletteer_d("ltc::trio4")

### problem 1: percent doesn't equal 100% 
  ### A: do position = "dodge"
### problem 2: try scale_y_log10() but HAVE to change zero values

combined_viz2 <- combined_viz %>%
  mutate(deviation = 1 - mean)

combined_viz2 <- combined_viz2 %>%
  mutate(log_dev = log1p(deviation * 10000))

combined_viz2 %>%
  ggplot(aes(x = prefix, y = log_dev, fill = source)) +
  geom_col(position = "dodge") +
  coord_cartesian(ylim = c(0, max(combined_viz2$log_dev, na.rm = TRUE))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
        legend.position = "right") +
  scale_fill_paletteer_d("ltc::trio4")


## heat map

combined_viz %>%
  ggplot(aes(x = source, y = prefix, fill = mean)) +
  geom_tile() +
  scale_fill_viridis_c(labels = scales::percent)
```



</details>



# Daikon Output

## read in daikon output

```{r}
daikon_output <- read.delim("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/daikon output/0_117_246_375_539.p.txt", header = FALSE)
```


## read in vcds

```{r}
files_daikon <- list.files(
  path = "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/daikon output/refrenced vcds",
  full.names = TRUE
)

vcd_daikon <- lapply(files_daikon, vcd2df_test, "long")
names(vcd_daikon) <- tools::file_path_sans_ext(basename(files_daikon))
list2env(vcd_daikon, envir = .GlobalEnv)
```

## compare files

```{r}
vcd_viz_file(M_AXI_ARVALID, "C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/daikon output/refrenced vcds")

vcd_viz_folder("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/daikon output/refrenced vcds")
# they're basically the same??
```

# work on Calvin recs 6/16

## interactive visualizations

```{r}
q <- vcd_viz_folder_int("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds_2")


df_viz <- four_vcd_df %>%
      mutate(n_timestamp = n_distinct(timestamp) * 4) %>%
      filter(value != 0) %>%
      group_by(register) %>%
      mutate(n_value_change = n(),
             activity = n_value_change / n_timestamp) %>%
      distinct(register, activity) %>%
      arrange(register)
    
    df_mean <- df_viz %>%
      mutate(prefix = sub("_.*", "", register)) %>%
      group_by(prefix) %>%
      summarise(
        mean = mean(activity),
        n_registers = n_distinct(register),
        min_activity = min(activity),
        max_activity = max(activity),
        median_activity = median(activity),
        .groups = "drop"
      )

p <- ggplot(df_mean, aes(x = prefix, y = mean, fill = prefix,
                               text = paste0(
                                 "Prefix: ", prefix, "\n",
                                 "# Registers: ", n_registers, "\n",
                                 "Min Activity: ", round(min_activity, 4), "\n",
                                 "Max Activity: ", round(max_activity, 4), "\n",
                                 "Median Activity: ", round(median_activity, 4)
                               ))) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 5),
      legend.position = "none"
    ) +
    labs(
    title = "Figure 3: Usage of Register Pathways for All Timestamps",
    subtitle = "Data from .vcd files ('alu_add_sub', 'alu_eq', 'alu_lts', 'alu_ltu')",
    x = "Register Prefix",
    y = "Percent of Total Usage")
  
  plotly::ggplotly(p, tooltip = "text")
  
ggplotly(q, tooltip = "text") %>%
  layout(
    annotations = list(
      list(
        text = "Data from .vcd file 'alu_add_sub'",
        x = 0,
        y = 1.035,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left"
      )
    )
  )


#vcd_viz_folder_int("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds")
```


## specifically shadow register

```{r}
w = vcd_viz_folder_int_shadow("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds_2", "prefix")

ggplotly(w, tooltip = "text") %>%
  layout(
    annotations = list(
      list(
        text = "Data from .vcd file 'alu_add_sub'",
        x = 0,
        y = 1.035,
        xref = "paper",
        yref = "paper",
        showarrow = FALSE,
        xanchor = "left"
      )
    )
  )


vcd_viz_folder_int_shadow("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela//daikon output/refrenced vcds", "prefix")
```



## 3 axis visualization

3d graphic - variance within the set (set being?)
   x = categorical/register family, 
   y = aggregated activity level, 
   z = std of activity levels within a family

### alter the data 

```{r}
test_1 = vcd2df_test("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/vcds/alu_add_sub.vcd", "long")
names(test_1)

tst1 = test_1 %>%
  mutate(subgroup = sub(".*_([^_]+).*", "\\1", register)) %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity, subgroup) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  mutate(mean = mean(activity),
         std = ifelse(n() > 1, sd(activity), 0)) %>%
  distinct(prefix, mean, std, subgroup)

names(tst1)


plot_ly(
  tst1,
  x = ~subgroup,
  y = ~mean,
  z = ~std,
  type = "scatter3d",
  mode = "markers+text",
  text = ~prefix,
  color = ~prefix) %>%
  layout(title = "basic viz.")

tst1_label <- tst1 %>%
  group_by(prefix) %>%
  mutate(label = if_else(row_number() == 1, prefix, NA_character_)) %>%
  ungroup()

plot_ly(
  tst1_label,
  x = ~subgroup,
  y = ~mean,
  z = ~std,
  type = "scatter3d",
  mode = "markers+text",
  text = ~label,
  color = ~prefix) %>%
  layout(title = "colors grouped + less labels")

# oh god this looks horrible
# what if you just specify a single subgroup to vizualize?

tst2 = test_1 %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  filter(grepl("^shadow_", register)) %>%
  group_by(register) %>%
  mutate(n_value_change = n(),
         activity = n_value_change / n_timestamp) %>%
  distinct(register, activity) %>%
  mutate(subgroup = sub(".*_([^_]+).*", "\\1", register)) %>%
  group_by(subgroup) %>%
  mutate(mean = mean(activity),
         std = ifelse(n() > 1, sd(activity), 0)) %>%
  distinct(subgroup, mean, std)


plot_ly(
  tst2,
  x = ~seq_along(subgroup),
  y = ~mean,
  z = ~std,
  type = "scatter3d",
  mode = "markers+text",
  text = ~subgroup,
  color = ~subgroup) %>%
  layout(title = "shadow reg. only")
```

### 2/23 Calvin Ideas for 3D

```{r, include = FALSE}
# can make interesting by:
  # keep only shadow BUT each one of the vcd files correspond to the shadow data related to a source signal 
  # we can probably plot more than one set of shadow values against it (and cut out non-interesting info)
  # plot the interesting families of registers from a different source register
  # dimensons: source signal (name of vcd file), some measure of interestingness (maybe 2 dims like size and color), register family name
    # what is interestingness? - higher std is more interesting, filter out std > 0.1 
    # TRY:  
        # dif color for each source signal (vcd file)? and then have register labels 
        # dif color for each register label and labels for vcd files
        # maybe try shape of dot for each thing?
        # also try doing a shaded polygon if registers tend to group
    # MAYBE:
        # graphs are divided by prefix for vcd files
```

### prefix 2 vcd

```{r}
# dimensons: source signal (name of vcd file), some measure of interestingness (maybe 2 dims like size and color), register family name

prefix_3d_viz_2vcd("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds_2")
```

### subgroup all vcd

```{r}
subgroup_3d_viz_all_vcd("C:/Users/marin/OneDrive/Desktop/Research Articles/Marianela/test_vcds")
```



### AI-generated ideas

#### 1) prefix × activity × consistency (std)

```{r}
library(dplyr)
library(plotly)

tst1 <- M_AXI_ARVALID %>%
  mutate(n_timestamp = n_distinct(timestamp)) %>%
  filter(value != 0) %>%
  group_by(register) %>%
  mutate(
    n_value_change = n(),
    activity = n_value_change / n_timestamp
  ) %>%
  distinct(register, activity) %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(prefix) %>%
  summarise(
    mean = mean(activity),
    std = ifelse(n() > 1, sd(activity), 0),
    n = n(),
    .groups = "drop"
  )



tst1 <- tst1 %>%
  mutate(prefix_id = as.numeric(factor(prefix)))

fig1 <- plot_ly(
  tst1,
  x = ~prefix_id,
  y = ~mean,
  z = ~std,
  type = "scatter3d",
  mode = "markers",
  
  color = ~prefix,
  colors = "Set2",
  
  size = ~n,
  sizes = c(5, 25),
  
  text = ~paste0(
    "Prefix: ", prefix,
    "<br>Mean: ", round(mean, 3),
    "<br>Std: ", round(std, 3),
    "<br>N: ", n),
  hoverinfo = "text")


fig1 <- fig1 %>%
  layout(
    scene = list(
      xaxis = list(title = "Prefix"),
      yaxis = list(title = "Mean activity"),
      zaxis = list(title = "Std deviation")
    )
  )

fig1

tst2 <- tst1 %>%
  mutate(
    prefix_id = as.numeric(factor(prefix)),
    prefix_group = paste0("Group ", ((prefix_id - 1) %/% 5) + 1)
  )

fig <- plot_ly(
  tst2,

  x = ~prefix_id,
  y = ~mean,
  z = ~std,

  type = "scatter3d",
  mode = "markers",

  # 🎨 darker grouped colors
  color = ~prefix_group,
  colors = "Dark2",

  # 🔺 inflate marker size
  size = ~(n * 10),
  sizes = c(80, 180),

  # 🧠 better hover info
  text = ~paste0(
    "Prefix: ", prefix,
    "<br>Group: ", prefix_group,
    "<br>Mean: ", round(mean, 3),
    "<br>Std: ", round(std, 3),
    "<br>N: ", n
  ),
  hoverinfo = "text"
)

fig
```

#### 2) activity surface plot

```{r}
# This shows where activity is concentrated over time.
# Higher peaks = more activity.

surface_df <- test_1 %>%
  filter(value != 0) %>%
  count(register, timestamp)

surface_mat <- surface_df %>%
  pivot_wider(
    names_from = timestamp,
    values_from = n,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("register") %>%
  as.matrix()

plot_ly(
  z = surface_mat,
  type = "surface"
)

plot_ly(
  x = colnames(surface_mat),
  y = rownames(surface_mat),
  z = surface_mat,
  type = "surface"
)
```

#### 3D Density Surface

```{r}
features <- test_1 %>%
  group_by(register) %>%
  summarize(
    mean = mean(value),
    std = sd(value),
    nonzero = sum(value != 0)
  )

density_df <- features %>%
  mutate(
    mean_bin = cut(mean, 25),
    std_bin = cut(std, 25)
  ) %>%
  filter(!is.na(mean_bin), !is.na(std_bin)) %>%
  count(mean_bin, std_bin)

density_mat <- density_df %>%
  tidyr::pivot_wider(
    names_from = std_bin,
    values_from = n,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("mean_bin") %>%
  as.matrix()

plot_ly(
  z = density_mat,
  type = "surface"
)
```

#### Network Graph of Register Similarity

```{r}
wide_df <- test_1 %>%
  pivot_wider(
    names_from = timestamp,
    values_from = value,
    values_fill = 0
  )

mat <- as.matrix(wide_df[,-1])
rownames(mat) <- wide_df$register

corr_mat <- cor(t(mat), use = "pairwise.complete.obs")
diag(corr_mat) <- 0
corr_mat[abs(corr_mat) < 0.8] <- 0

edges <- which(corr_mat != 0, arr.ind = TRUE)

edge_df <- data.frame(
  from = rownames(corr_mat)[edges[,1]],
  to   = colnames(corr_mat)[edges[,2]],
  weight = corr_mat[edges]
)

# simple circular layout (no igraph needed)
nodes <- unique(c(edge_df$from, edge_df$to))
n <- length(nodes)

node_df <- data.frame(
  node = nodes,
  x = cos(2 * pi * seq_len(n) / n),
  y = sin(2 * pi * seq_len(n) / n)
)

edge_df <- edge_df %>%
  left_join(node_df, by = c("from" = "node")) %>%
  rename(x0 = x, y0 = y) %>%
  left_join(node_df, by = c("to" = "node")) %>%
  rename(x1 = x, y1 = y)

plot_ly() %>%
  add_segments(
    data = edge_df,
    x = ~x0, xend = ~x1,
    y = ~y0, yend = ~y1,
    line = list(width = 1)
  ) %>%
  add_markers(
    data = node_df,
    x = ~x, y = ~y,
    text = ~node,
    hoverinfo = "text",
    size = 8
  )
```

#### 3D Bubble Plot

```{r}
features <- test_1 %>%
  group_by(register) %>%
  summarize(
    mean = mean(value),
    std = sd(value),
    nonzero = sum(value != 0)
  )

plot_ly(
  features,
  x = ~mean,
  y = ~std,
  z = ~nonzero,
  size = ~nonzero,
  color = ~nonzero,
  type = "scatter3d",
  mode = "markers"
)
```

#### Functional Cluster Hulls

```{r}
features <- test_1 %>%
  mutate(prefix = sub("_.*", "", register)) %>%
  group_by(register, prefix) %>%
  summarize(
    mean = mean(value),
    std = sd(value),
    nonzero = sum(value != 0),
    .groups = "drop"
  )

plot_ly(
  features,
  x = ~mean,
  y = ~std,
  z = ~nonzero,
  color = ~prefix,
  type = "scatter3d",
  mode = "markers"
)
```

#### PCA Embedding

```{r}
wide_df <- test_1 %>%
  pivot_wider(
    names_from = timestamp,
    values_from = value,
    values_fill = 0
  )

mat <- as.matrix(wide_df[,-1])

# 1. keep only active signals
activity <- apply(mat, 2, function(x) mean(x != 0))

mat <- mat[, activity > 0.01]

# 2. fill gaps
mat[is.na(mat)] <- 0

# 3. remove dead signals
mat <- mat[, apply(mat, 2, var) > 0]

pca <- prcomp(
  mat,
  scale. = FALSE
)

pca_df <- as.data.frame(
  pca$x[,1:3]
)

pca_df$register <- wide_df$register
pca_df$prefix <- sub("_.*", "", wide_df$register)

plot_ly(
  pca_df,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~prefix,
  text = ~register,
  type = "scatter3d",
  mode = "markers"
)
```



